Meta-analytic approaches and new effect sizes for treatment effects in comparative physiology

### 2021-05-17 MS/ms.Rmd
just updating the outline a bit, few wording change sto make things a little more clear / accurate

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Distilling published research results into standardised effect sizes that can be weighted by their sample size (i.e., inverse sampling variance) and compared across broad sets of research designs, study systems and species is its core objective and strength. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing “apples with oranges”. Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-analytic models to both account for factors driving differences across studies and make them more comparable. In addition, we derive new effect size measures that provide comparative physiologists with a means to directly make effect sizes comparable without the need to resort to more complex statistical models that may be harder to interpret. Our ‘new’ effect sizes and corresponding sampling variances help physiologists deal with temperature and dosage differences across studies; allowing researchers to compare both mean differences (e.g., Q10 to compare mean differences in physiological rates over 10 degrees) and associated differences in variance (e.g, SQ10 for comparing variability in in physiological rates over 10 degrees) for physiological traits. The new effect sizes we propose, in combination with existing meta-analytic models, will pave the way for comparative physiologists to explore exciting new questions by making results from large-scale data sets from the literature more accessible and widely interpretable.

@@

@@

>>>>

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Distilling published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance) in a meta-analysis, and compared across broad sets of research designs, study systems and species is its core objective and strength. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing “apples with oranges”. Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-analytic models to both account for factors driving differences across studies and make them more comparable. In addition, we derive new effect size measures that provide comparative physiologists with a means to directly make effect sizes comparable without the need to resort to more complex statistical models that may be harder to interpret. Our ‘new’ effect sizes and corresponding sampling variances help physiologists deal with common treatment differeces across studies; allowing researchers to compare both mean differences (e.g., Q10 to compare mean differences in physiological rates over 10 degrees) and associated differences in variance (e.g, SQ10 for comparing variability in in physiological rates over 10 degrees) for physiological traits. The new effect sizes we propose, in combination with existing meta-analytic models, will pave the way for comparative physiologists to explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

## **Survey of Meta-Analytic Practices in Comparative Physiology**

#### *Literature Search and Data Collection*
To understand meta-analytic practices in comparative physiology, we searched for meta-analyses tackling comparative physiological questions that relied on data compiled from studies using animal model systems with more than a single species. All searches were conducted......**write details here**.

We excluded plant meta-analyses because they were often focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output and greenhouse gas emissions) that did not focus on organism physiological function across species. Meta-analyses needed to have collected physiological trait data from the literature including traits, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (heart rates, sprint speed, swim speed, growth rates) and immune traits (cell function, antibodies, ect). Studies were also included if they measured reproductive allocation, survival / mortality or quantified scaling relationships with body size (e.g., metabolic scaling) across taxa. Studies on behavior were only included if they were explicitly interested in proximate physiological drivers of behavioral variation. Meta-analyses quantifying phenology, differences in body sizes between the sexes (i.e., SSD) and those focused on validating physiological methodology were excluded because the underlying question was not physiological in nature. 
#### *Current Meta-analytic Practices in Comparative Physiology*

## **Opening up New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes and Sampling Variances**

#### *Brief Introduction to Common Effect Sizes and Sampling Errors that are Useful in Comparative Physiology*
#### *Expanding the Scope of Effect Size Metrics in Comparative Physiology*
#### *New Effect Size Metrics that Account for Variability in Treatments*

@@

#### *Multi-level Meta-Analysis when the Overall Effect is of Interest*
#### *Quantifying Heterogenity in Effects: Is it as variable as you think?*
#### *Multi-level Metaregression to Understand Variation and Imrpove Interpretation*

## **Conclusions and Future Directions**
